---
title: "Paper outline"
subtitle: "Testing LD clusters instead of SNPs --- structure, evidence inventory, and gaps"
author: "module_sim_LDscnR"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  pdf_document:
    number_sections: true
    toc: true
    toc_depth: 2
    latex_engine: pdflatex
geometry: margin=2.4cm
fontsize: 10pt
header-includes:
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage[table]{xcolor}
  - \definecolor{shade}{HTML}{F2F5F6}
  - \definecolor{gapc}{HTML}{FBF3EE}
  - \usepackage{mdframed}
  - \newmdenv[backgroundcolor=shade,linewidth=0pt,skipabove=8pt,skipbelow=8pt]{keybox}
  - \newmdenv[backgroundcolor=gapc,linewidth=0pt,skipabove=8pt,skipbelow=8pt]{gapbox}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
suppressMessages({library(knitr)})
```

# The argument in one paragraph

A genome scan tests SNPs; the object of interest is a locus. Benjamini--Hochberg
controls the false discovery proportion **among the units tested**, so a
single-SNP scan controls it among SNPs while being judged against a locus-level
truth. On one simulated panel, 2,467 significant SNPs were **132 distinct LD
clusters**, with the rejected set 6.3 times more redundant than the genome as a
whole --- so "5% FDR over 2,467 discoveries" is a guarantee about a different
object from the one being counted. Testing near-independent LD clusters aligns
the two. The paper is that argument plus a pipeline that derives its units from
the data's own LD decay rather than from chosen constants.

\begin{keybox}
\textbf{The central claim requires no simulation.} It follows from what BH
controls. Everything measured below is supporting evidence for a claim that
stands on its own, which is an unusual and exploitable position: the simulations
show the size of the effect, not whether it exists.
\end{keybox}

# Why it is one paper and not four

Each link is load-bearing rather than merely sequential. Remove any one and the
argument fails at a specific point.

```{r chain, results="asis"}
cat("\\begin{tabular}{@{}p{3.4cm}p{10.8cm}@{}}\n\\toprule\n")
cat("\\textbf{Link} & \\textbf{What fails without it} \\\\\n\\midrule\n")
rows <- list(
 c("LD-decay estimation","The clustering thresholds become chosen constants. The decay fit derives \\texttt{min\\_r2} and the merge distance \\emph{per chromosome}, so a threshold means the same thing on chromosomes with different recombination."),
 c("$ld_w$","No units. $ld_w$ is what makes a cluster a cluster --- and, separately, is shown \\emph{not} to work as a filter within them, a negative that earns its place by ruling out the obvious alternative use."),
 c("Two-stage reduction","The reported FDR is over SNPs. This is the claim."),
 c("Association test","Nothing --- and that is the point. EMMAX and LFMM give the same conclusions, which is what shows the result is a property of the units rather than of any test."),
 c("Consistency analysis","No guidance on which derived thresholds the answer depends on, so the pipeline is not usable by anyone else."))
for (r in rows) cat(paste(r, collapse=" & "), "\\\\\n")
cat("\\bottomrule\n\\end{tabular}\n")
```

# Proposed structure

## 1. Introduction

The redundancy problem, stated as an FDR-target mismatch rather than as a
multiple-testing complaint. Existing LD-aware scans mostly prune or clump *after*
testing; this reverses the order and makes the correction match the unit.

## 2. Methods

2.1 Decay estimation and the derived thresholds. 2.2 $ld_w$. 2.3 Two-stage
complexity reduction. 2.4 Cluster-level testing --- **Simes over members**, which
beat every single-marker summary tested. 2.5 The consistency analysis.

## 3. Results

```{r res, results="asis"}
cat("\\begin{tabular}{@{}p{4.4cm}p{6.4cm}p{3.2cm}@{}}\n\\toprule\n")
cat("\\textbf{Section} & \\textbf{Headline} & \\textbf{Status} \\\\\n\\midrule\n")
rows <- list(
 c("3.1 The redundancy problem","2,467 significant SNPs are 132 clusters; small-$p$ tail 6.3$\\times$ more redundant than the genome","\\textbf{done}"),
 c("3.2 What clustering buys","Precision 2$\\times$ higher (62 of 71 panels). A median 51 of 67 single-SNP flags tag no QTN, against 7 of 10","\\textbf{done}"),
 c("3.3 How to summarise a cluster","Simes $>$ eMLG $>$ representative $>$ min-$p$ $>$ highest-$ld_w$","\\textbf{done}"),
 c("3.4 The metric is a claim","$F_\\beta$ crossover at $\\beta \\approx 2$; below it clustering wins, above it does not","\\textbf{done}"),
 c("3.5 Engine replication","EMMAX and LFMM: same composition result, same crossover","\\textbf{done}"),
 c("3.6 Sensitivity","Two axes --- selection (27-cell grid) and test (engine) --- reported separately","\\textbf{not run}"),
 c("3.7 Empirical demonstration","---","\\textbf{missing}"))
for (r in rows) cat(paste(r, collapse=" & "), "\\\\\n")
cat("\\bottomrule\n\\end{tabular}\n")
```

## 4. What does not work

Short, and it earns its place by ruling out the obvious things a reader will
propose. The C-score as a ranking statistic; four filters ($ld_w$ top-$k$, MAF,
cluster size, geometric exclusion) all strongly enriched for QTN-bearing units
and none improving precision-recall. **Enrichment predicts precision within a
restriction, but no restriction beats using every cluster** --- the recall lost
always exceeds the precision gained.

## 5. Limitations


# Parameters that transfer, and the currency each transfers in

The pipeline has three tunable thresholds. All three were being quoted as
*numbers*, and measuring what each number actually does --- across the eight
simulation settings, and against the stickleback panel --- shows that **each
transfers in a different currency, and for two of them the naive choice is
wrong.**

```{r transfer, results="asis"}
cat("\\begin{tabular}{@{}p{2.5cm}p{4.1cm}p{3.5cm}p{4.0cm}@{}}\n\\toprule\n")
cat("\\textbf{Parameter} & \\textbf{Across sim settings} & \\textbf{To the panel} & \\textbf{Transfers as} \\\\\n\\midrule\n")
cat("\\texttt{distance\\_} \\texttt{threshold} 1e5 & \\textbf{fails}: 0.60--10.94\\% of flagged-cluster gaps exceed it, an 18$\\times$ spread; switching to the derived value removes \\textbf{68--100\\%} of split points in every setting & \\textbf{also consequential}, in the opposite direction: derived is 3--18 kb, \\emph{below} the hand-set value & an \\textbf{operating point} (gap exceedance), never a bp value --- but the operating point itself does not transfer \\\\\n\\addlinespace\n")
cat("size floor 2 & \\textbf{stable}: retains 46.0--51.2\\% of clusters across all eight settings, a 1.1$\\times$ spread & \\textbf{does not}: 17.0\\%, a 2.8$\\times$ gap & a \\textbf{structural value} --- Simes over a singleton \\emph{is} that marker's $p$ --- with the differing retention reported as a result \\\\\n\\addlinespace\n")
cat("\\texttt{ld\\_w\\_} \\texttt{threshold} 0.025 & \\textbf{fails}: retains 5.1\\% to 48.1\\% of markers, a 9.4$\\times$ spread & panel's 28\\% sits \\emph{inside} the simulation range & \\textbf{marker retention}, matched per dataset; the numeric value carries no meaning \\\\\n")
cat("\\bottomrule\n\\end{tabular}\n")
```

\begin{keybox}
\textbf{Only the size floor transfers, and pooling hid both failures.} Averaged
over cells, \texttt{ld\_w} retention reads 55/20/5\% against the panel's
54/28/13\% --- close enough to call the axis well matched, which is what was
reported to the panel session. Broken out, that 20\% is the average of 5.1\% and
48.1\%.

\textbf{The two failing axes are also not independent.} Gaps exceeding the
distance threshold track the number of flagged stage-1 clusters (Spearman
$-0.905$), and \texttt{ld\_w\_threshold} is what sets that count --- 1,272 to
32,318 across the eight settings. So a factorial grid cannot vary these two
orthogonally and read the margins separately, which is a constraint on the
sensitivity design rather than a caveat on it.
\end{keybox}

\begin{keybox}
\textbf{A withdrawn claim, and how it was caught.} This table first read
"\texttt{distance\_threshold} inert everywhere, derived exceeds the largest gap
on 160 of 160 chromosomes". That was computed on adjacent-marker gaps over the
whole map; \texttt{split\_by\_distance()} uses gaps between consecutive
\emph{flagged stage-1 clusters}. Wrong unit, wrong subset, and the correct
quantity reverses the conclusion. Found by the panel session reading the source
rather than by any disagreement in the numbers --- which is the second time
today a result broke under decomposition rather than under challenge.
\end{keybox}

\noindent The practical rule that follows is **different invariants for
different axes** --- match marker retention on `ld_w`, hold the size floor at an
identical value and report the differing retention, and specify
`distance_threshold` by what fraction of gaps it splits rather than in bp. A
grid whose three axes all read as matched numbers would be comparing three
different things.

## Why this is a result and not housekeeping

The `distance_threshold` case is the sharpest. The package derives it as
$d = \rho/(a(1-\rho))$, so the window **collapses as decay speeds up** and
carries no term for the size of the feature stage-2 merging exists to bridge. On
the panel that yields a 6 kb median merge distance on a genome whose motivating
feature is a 430 kb inversion. Here it yields roughly 1 Mb --- which is *not*
past the top of the flagged-cluster gap distribution, and adopting it would
remove 68--100% of split points depending on the setting. Identical rule,
consequential on both datasets, and in opposite directions: too small to span
the features there, and physically meaningless in its tail here (4,961 Mb
derived on a 30 Mb chromosome). Three fixes were proposed and all three refuted,
each acting hardest exactly where the local estimate is most informative.

# What still needs to be done {#sec:todo}

Ordered by what blocks what.

```{r todo, results="asis"}
cat("\\begin{tabular}{@{}p{0.4cm}p{5.2cm}p{8.4cm}@{}}\n\\toprule\n")
rows <- list(
 c("1","\\textbf{Check whether the cells are comparable at all}","\\texttt{ld\\_w\\_threshold} 0.025 retains 5--6\\% of markers in \\texttt{V0.5\\_c1} and 26--48\\% in \\texttt{V0.5\\_c2}. The cell where clustering wins 19/19 is at the most aggressive end. This does not invalidate anything --- unflagged markers are still clustered and tested --- but the cross-cell comparison carrying the headline result has not been checked at matched operating points. \\textbf{Newly surfaced; do this before the sensitivity grid.}"),
 c("2","\\textbf{The factorial sensitivity grid}","$3\\times3\\times3$, 9 re-clusterings. Now better specified than when it was planned: report \\emph{operating points} rather than values, never pool across cells, and report the two axes (selection vs test) separately."),
 c("3","$T_2$ / $T_2$b for the local threshold","Scoped down: since the local rule collapses to a flat threshold on the panel, the most $T_2$ can support is \\emph{helps in the cold-block regime, inert to degenerate outside it}. Worth deciding whether that claim is wanted before spending the time."),
 c("4","\\textbf{An empirical demonstration}","Unchanged and still the largest gap. See below."),
 c("5","\\textbf{Decide what to do about the hardcoded 1e5}","It appears in roughly 20 scripts. It is \\emph{not} inert -- that claim was withdrawn -- so this is a live choice rather than documentation: adopting the derived rule would remove 68--100\\% of split points and repartition every result. Nothing needs re-running to keep the current results valid, but the choice has to be argued rather than inherited."))
for (r in rows) cat(paste(r, collapse=" & "), "\\\\\n\\addlinespace\n")
cat("\\bottomrule\n\\end{tabular}\n")
```

## Settled since the first draft of this outline

- **`n_win_decay` = 20**, by two independent criteria --- an internal consistency
  measure and an external one scored against the simulation's true recombination
  rate. Not 50, which buys nothing for 2.5 times the windows.
- **The recombination proxy is real**: Spearman +0.757 against exact truth, 20 of
  20 chromosomes. The operational figure is **0.782** at $w = 20$, not the 0.830
  headline, which was a composite of cold-block *detection* and continuous
  *tracking*.
- **The cap is generous or nothing**: the cost is monotone, so there is no
  optimum; 10 Mb costs 0.004 and bounds a quantity that otherwise reaches
  4,961 Mb.
- **The size floor is inert above 2** ($p = 0.11$, 0.40, 0.053 for floor 2 to 8),
  and floor 1 to 2 *gains* recall while *losing* precision --- so floor 2 keeps
  its structural justification but not the performance gloss.
- **The panel can falsify the proxy but cannot calibrate it**, because
  calibration needs variation in the cold-block regime it lacks. A property of
  the stickleback map, not of empirical data in general.

# Evidence inventory: what would a referee ask for?

\begin{gapbox}
\textbf{Gap 1 --- no empirical demonstration.} Everything is simulated. The
natural companion is the 3sp panel, but its support statistic currently measures
the \emph{filter} rather than the test (units the filter selected but the test did
not reject overlap peaks at 69.2\% against the discoveries' 78.2\%, $p = 0.18$),
so it cannot carry a validation claim as it stands. \textbf{This is the largest
gap.}
\end{gapbox}

\begin{gapbox}
\textbf{Gap 2 --- the advantage is concentrated in one simulation cell.} Simes
beats the single-SNP scan in 19 of 19 panels in \texttt{V0.5\_c1} and the paired
tests in the other three cells are \textbf{null, not adverse} (26 W / 22 L, all
$p > 0.4$) --- several panels are ties in which neither analysis recovers
anything. Must be stated in the Results, not a supplement: \emph{decisive where
there is signal, indistinguishable where there is not.}
\end{gapbox}

\begin{gapbox}
\textbf{Gap 3 --- the sensitivity analysis is planned, not run}, and it now has
a prerequisite it did not have: the four cells sit at very different
\texttt{ld\_w} operating points (item 1, \S\ref{sec:todo}), so a grid run over
them would inherit that. Design is in \texttt{TODO\_sensitivity.md}, amended to
report operating points rather than values and never to pool across cells.
\end{gapbox}

\begin{gapbox}
\textbf{Gap 4 --- one simulation generation.} Re-clustering at the newer decay
settings changes nothing ($p = 0.25$ and $0.85$), so this is closed for the decay
parameter --- but all results come from one simulator under one demographic
model.
\end{gapbox}

# Claims the paper should not make

- That clustering finds more loci. It finds **fewer** and is right more often.
- That $ld_w$ filtering helps. Four variants, none did.
- That any performance number transfers. The **ratios** are the transferable
  part; absolute precision is a property of this signal-to-noise regime.
- That a threshold is recommended for the consistency score. Every threshold
  tried on a C-like quantity failed, including the one that looked principled.
- That enrichment for containing a QTN is irrelevant to detection. It predicts
  precision; it just cannot be exploited by restriction.
- That a threshold means the same thing on two datasets because it is the same
  number. Measured, one of the three transfers, one does not, and one fails even
  across simulation settings.
- That raising the minimum cluster size improves precision. Measured, floor 1 to
  2 *lowers* it.

# Two methodological findings that may be worth their own space

\begin{keybox}
\textbf{Name the axis a stability measure is on.} A consistency score computed
over parameters that move only the selection measures how stable the
\emph{selection} is. "Called under every setting" reads as a claim about the
finding. The second association engine is the missing test-side axis, which makes
the engine sweep and the sensitivity analysis one design rather than two.

\textbf{Replicate specifically where nobody doubts the result.} The LFMM
replication on the panel side surfaced two sampler defects that had been sitting
in the original arm for a day with a manuscript being written around them. A
replication that \emph{agrees} leaves such defects in place; the disagreement is
the diagnostic.
\end{keybox}

# Provenance

Result document: `ld_clustering_result.pdf`. Full record including everything
withdrawn: `development_history.pdf`. Both build from committed CSVs. Branch
`null-rework-two-basis`.
