---
title: "Paper outline"
subtitle: "Testing LD clusters instead of SNPs --- structure, evidence inventory, and gaps"
author: "module_sim_LDscnR"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  pdf_document:
    number_sections: true
    toc: true
    toc_depth: 2
    latex_engine: pdflatex
geometry: margin=2.4cm
fontsize: 10pt
header-includes:
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage[table]{xcolor}
  - \definecolor{shade}{HTML}{F2F5F6}
  - \definecolor{gapc}{HTML}{FBF3EE}
  - \usepackage{mdframed}
  - \newmdenv[backgroundcolor=shade,linewidth=0pt,skipabove=8pt,skipbelow=8pt]{keybox}
  - \newmdenv[backgroundcolor=gapc,linewidth=0pt,skipabove=8pt,skipbelow=8pt]{gapbox}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
suppressMessages({library(knitr)})
```

# The argument in one paragraph

A genome scan tests SNPs; the object of interest is a locus. Benjamini--Hochberg
controls the false discovery proportion **among the units tested**, so a
single-SNP scan controls it among SNPs while being judged against a locus-level
truth. On one simulated panel, 2,467 significant SNPs were **132 distinct LD
clusters**, with the rejected set 6.3 times more redundant than the genome as a
whole --- so "5% FDR over 2,467 discoveries" is a guarantee about a different
object from the one being counted. Testing near-independent LD clusters aligns
the two. The paper is that argument plus a pipeline that derives its units from
the data's own LD decay rather than from chosen constants.

\begin{keybox}
\textbf{The central claim requires no simulation.} It follows from what BH
controls. Everything measured below is supporting evidence for a claim that
stands on its own, which is an unusual and exploitable position: the simulations
show the size of the effect, not whether it exists.
\end{keybox}

# Why it is one paper and not four

Each link is load-bearing rather than merely sequential. Remove any one and the
argument fails at a specific point.

```{r chain, results="asis"}
cat("\\begin{tabular}{@{}p{3.4cm}p{10.8cm}@{}}\n\\toprule\n")
cat("\\textbf{Link} & \\textbf{What fails without it} \\\\\n\\midrule\n")
rows <- list(
 c("LD-decay estimation","The clustering thresholds become chosen constants. The decay fit derives \\texttt{min\\_r2} and the merge distance \\emph{per chromosome}, so a threshold means the same thing on chromosomes with different recombination."),
 c("$ld_w$","No units. $ld_w$ is what makes a cluster a cluster --- and, separately, is shown \\emph{not} to work as a filter within them, a negative that earns its place by ruling out the obvious alternative use."),
 c("Two-stage reduction","The reported FDR is over SNPs. This is the claim."),
 c("Association test","Nothing --- and that is the point. EMMAX and LFMM give the same conclusions, which is what shows the result is a property of the units rather than of any test."),
 c("Consistency analysis","No guidance on which derived thresholds the answer depends on, so the pipeline is not usable by anyone else."))
for (r in rows) cat(paste(r, collapse=" & "), "\\\\\n")
cat("\\bottomrule\n\\end{tabular}\n")
```

# Proposed structure

## 1. Introduction

The redundancy problem, stated as an FDR-target mismatch rather than as a
multiple-testing complaint. Existing LD-aware scans mostly prune or clump *after*
testing; this reverses the order and makes the correction match the unit.

## 2. Methods

2.1 Decay estimation and the derived thresholds. 2.2 $ld_w$. 2.3 Two-stage
complexity reduction. 2.4 Cluster-level testing --- **Simes over members**, which
beat every single-marker summary tested. 2.5 The consistency analysis.

## 3. Results

```{r res, results="asis"}
cat("\\begin{tabular}{@{}p{4.4cm}p{6.4cm}p{3.2cm}@{}}\n\\toprule\n")
cat("\\textbf{Section} & \\textbf{Headline} & \\textbf{Status} \\\\\n\\midrule\n")
rows <- list(
 c("3.1 The redundancy problem","2,467 significant SNPs are 132 clusters; small-$p$ tail 6.3$\\times$ more redundant than the genome","\\textbf{done}"),
 c("3.2 What clustering buys","Precision 2$\\times$ higher (62 of 71 panels). A median 51 of 67 single-SNP flags tag no QTN, against 7 of 10","\\textbf{done}"),
 c("3.3 How to summarise a cluster","Simes $>$ eMLG $>$ representative $>$ min-$p$ $>$ highest-$ld_w$","\\textbf{done}"),
 c("3.4 The metric is a claim","$F_\\beta$ crossover at $\\beta \\approx 2$; below it clustering wins, above it does not","\\textbf{done}"),
 c("3.5 Engine replication","EMMAX and LFMM: same composition result, same crossover","\\textbf{done}"),
 c("3.6 Sensitivity","Two axes --- selection (27-cell grid) and test (engine) --- reported separately","\\textbf{not run}"),
 c("3.7 Empirical demonstration","---","\\textbf{missing}"))
for (r in rows) cat(paste(r, collapse=" & "), "\\\\\n")
cat("\\bottomrule\n\\end{tabular}\n")
```

## 4. What does not work

Short, and it earns its place by ruling out the obvious things a reader will
propose. The C-score as a ranking statistic; four filters ($ld_w$ top-$k$, MAF,
cluster size, geometric exclusion) all strongly enriched for QTN-bearing units
and none improving precision-recall. **Enrichment predicts precision within a
restriction, but no restriction beats using every cluster** --- the recall lost
always exceeds the precision gained.

## 5. Limitations

# Evidence inventory: what would a referee ask for?

\begin{gapbox}
\textbf{Gap 1 --- no empirical demonstration.} Everything is simulated. The
natural companion is the 3sp panel, but its support statistic currently measures
the \emph{filter} rather than the test (units the filter selected but the test did
not reject overlap peaks at 69.2\% against the discoveries' 78.2\%, $p = 0.18$),
so it cannot carry a validation claim as it stands. \textbf{This is the largest
gap.}
\end{gapbox}

\begin{gapbox}
\textbf{Gap 2 --- the advantage is concentrated in one simulation cell.} Simes
beats the single-SNP scan in 19 of 19 panels in \texttt{V0.5\_c1} and the paired
tests in the other three cells are \textbf{null, not adverse} (26 W / 22 L, all
$p > 0.4$) --- several panels are ties in which neither analysis recovers
anything. Must be stated in the Results, not a supplement: \emph{decisive where
there is signal, indistinguishable where there is not.}
\end{gapbox}

\begin{gapbox}
\textbf{Gap 3 --- the sensitivity analysis is planned, not run.} It is also the
link that makes the pipeline usable by someone else. Design is in
\texttt{TODO\_sensitivity.md}: $3\times3\times3$ over \texttt{ld\_w\_threshold},
\texttt{min\_r2\_rho} and minimum cluster size (9 re-clusterings, not 27, since
the size floor is a post-filter), reported on two axes with no threshold
recommended.
\end{gapbox}

\begin{gapbox}
\textbf{Gap 4 --- one simulation generation.} Re-clustering at the newer decay
settings changes nothing ($p = 0.25$ and $0.85$), so this is closed for the decay
parameter --- but all results come from one simulator under one demographic
model.
\end{gapbox}

# Claims the paper should not make

- That clustering finds more loci. It finds **fewer** and is right more often.
- That $ld_w$ filtering helps. Four variants, none did.
- That any performance number transfers. The **ratios** are the transferable
  part; absolute precision is a property of this signal-to-noise regime.
- That a threshold is recommended for the consistency score. Every threshold
  tried on a C-like quantity failed, including the one that looked principled.
- That enrichment for containing a QTN is irrelevant to detection. It predicts
  precision; it just cannot be exploited by restriction.

# Two methodological findings that may be worth their own space

\begin{keybox}
\textbf{Name the axis a stability measure is on.} A consistency score computed
over parameters that move only the selection measures how stable the
\emph{selection} is. "Called under every setting" reads as a claim about the
finding. The second association engine is the missing test-side axis, which makes
the engine sweep and the sensitivity analysis one design rather than two.

\textbf{Replicate specifically where nobody doubts the result.} The LFMM
replication on the panel side surfaced two sampler defects that had been sitting
in the original arm for a day with a manuscript being written around them. A
replication that \emph{agrees} leaves such defects in place; the disagreement is
the diagnostic.
\end{keybox}

# Provenance

Result document: `ld_clustering_result.pdf`. Full record including everything
withdrawn: `development_history.pdf`. Both build from committed CSVs. Branch
`null-rework-two-basis`.
